---
title: "Using Dynamic TOPMODEL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Dynamic TOPMODEL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The purpose of this vignette is to provide an outline of the steps needed to
build a dynamic TOPMODEL implementation using the dynatopGIS package.

# Implementation notes

The dynatopGIS package implements a structured, object orientated, data
flow. The steps outlined below create a `dynatopGIS` catchment object to which
actions are then applied to generate a model. 

The `dynatopGIS` package is written using the object orientated framework
provided by the `R6` package. This means that some aspects of working with the
objects may appear idiosyncratic for some R users. In using the package as
outlined in this vignette these problems are largely obscured, except for the
call structure. However, before adapting the code, or doing more complex analysis
users should read about `R6` class objects (e.g. in the `R6` package vignettes
or in the Advanced R book). One particular gotcha is when copying an object. Using

```{r eval=FALSE}
my_new_object  <-  my_object
```

creates a pointer, that is altering `my_new_object` also alters
`my_object`. To create a new independent copy of `my_object` use

```{r eval=FALSE}
my_new_object  <-  my_object$clone()
```

# Getting started

The dynatopGIS packages works through a number of steps to generate a Dynamic
TOPMODEL object suitable for use in with the `dynatop` package. Each step
generates one or more layers which are saved as raster or shape files into the projects
working directory (which is not necessarily the R working directory). A record
of these layers is kept in the json format meta data file.

This vignette demonstrates the use of the `dynatopGIS` package using data from the Swindale
catchment in the UK.

To start first load the library

```{r load_library}
library("dynatop")
```

For this vignette we will store the data into a temporary directory

```{r tempory_dir}
demo_dir <- tempdir()
```
and initialise the analysis by creating a new object specifying the location
of the meta data file, which will be created if it doesn't exist.

```{r, initialisation}
ctch <- dynatopGIS$new(file.path(demo_dir,"example.tif"))
```

# Adding catchment data

The basis of the analysis is a rasterised Digital Elevation Model (DEM) of
the catchment and a vectorised representation of the river network with
attributes. Currently these can be in any format supported by the ```terra```
library.

However, within the calculations used for sink filling,
flow routing and topographic index calculations the raster DEM is presumed
to be projected so that is has square cells such that
the difference between the cell centres (in meters) does not alter.


For Swindale the suitable DEM and channel files can be found using:

```{r, data_files}
dem_file <- system.file("extdata", "gis", "SwindaleDTM40m.tif", package="dynatop", mustWork = TRUE)
channel_file <- system.file("extdata", "gis", "SwindaleRiverNetwork.shp", package="dynatop", mustWork = TRUE)
```

Before adding either the DEM or channel a raster map of the catchment outline must be
provided. This defines not only the catchment boundaries but also, if
required, subcatchments, each of which must be given a unique number. The
projection and resolution of this map is used in all subsequent GIS processing

In this example the catchment map is generated from the DEM, which, by
convention must contain a edge rows and columns containing only `NA` values.

```{r, add_catchment}
dem <- terra::rast(dem_file)
dem <- terra::extend(dem,1) ## pad with NA values
catchment_outline <- terra::ifel(is.finite(dem),1,NA)
ctch$add_catchment(catchment_outline)
```

Either the DEM or channel files can be added to the project first. In this
case we add the DEM with

```{r, add_dem}
ctch$add_dem(dem)
```

Adding river channel data is more complex. The `add_channel` method requires a
`SpatVector` object (or a file name that can be loaded as a `SpatVect`
object).
Each vector object is treated as a length of river channel
which requires the following properties

-   *name* - a label for the channel length
-   *endNode* - a label for the downstream end of the river length
-   *startNode* - a label for the upstream end of the river length
-   *length* - the length in meters
-   *area* - surface area in square meters
-   *width* - width of the channel
-   *slope* - bed slope of the channel

Additional properties are currently kept but ignored with the exception of
*id* which is overwritten.
	
Since it is possible that these properties are present in a data file
under different names some basic preprocessing may be required.
The `convert_channel` function is designed to help with this. 
To illustrate this let us examine the river network for Swindale

```{r, channel_current}
sp_lines <- terra::vect(channel_file)
head(sp_lines)
```

Some of the main properties are present under appropriate names (startNode,
endNode, length) but the remainder are missing. Also the river network is
defined as a series of lines, rather then polygons. The `convert_channel`
function addresses these shortcomings by 
- changing the names of the required properties
- buffering the line objects to create polygons

The `convert_channel` function takes a named vector giving the variable names
to be use for the properties. If we want to carry over the identifier as the name we could call
`convert_channel` with as follows:

```{r, channel_properties}
property_names <- c(name="identifier",
                    endNode="endNode",
                    startNode="startNode",
                    length="length")
chn <- convert_channel(sp_lines,property_names)
```
Since the data set for Swindale does not contain a channel width or slope default
values are used and warnings issued.

The river network can then be added to the project by

```{r, add_channel}
ctch$add_channel(chn)
```

# Getting and plotting catchment information

The `dynatopGIS` class has methods for returning and plotting the GIS data in
the project. The names of all the different GIS layers stored is returned by

```{r, list_layers}
ctch$get_layer()
```

These can be plotted (with or without the channel), for example

```{r, plot}
ctch$plot_layer("dem", add_channel=TRUE)
```

or returned, for example

```{r, get_layer}
ctch$get_layer("dem")
```

All layers are returned as `SpatRast` objects with the exception of the
`channel_vect` layer which is returned as a `SpatVect` object. 


# Filling sinks

For the hill slope to be connected to the river network all DEM cells must
drain to those that intersect with the river network. The algorithm
implemented in the `sink_fill` method ensures this is the case. 
In calling the `sink_fill` method a flow direction algorithm
is specified and the resulting flow paths recorded. If subcatchments are
present in the catchment map then only flow paths within the subcatchment are
considered.

The algorithm of used by the `sink_fill` method is iterative and the execution time of the
function is limited by capping the maximum number of iterations. If
this limit is reached without completion the method can call again with the
"hot start" option to continue from where it finished.

For Swindale, where the example DEM is already partially filled the algorithm
only alters a small area near the foot of the catchment.

```{r, sink_fill}
ctch$sink_fill()

terra::plot( ctch$get_layer('filled_dem') - ctch$get_layer('dem'),
            main="Changes to height")
```

# Computing properties

** THIS IS INCORRECT NO PROPERTIES ARE NEEDED IN dynatop since no grouping
into HRUs is used **
Two sets of properties are required for Dynamic TOPMODEL. The first set is
those required within the evaluation of the model; gradient and contour
length. The second set are those used for dividing the catchment up into
Hydrological Response Units (HRUs). Traditionally the summary used for the
separation of the HRUs is the topographic index, which is the natural
logarithm of the upslope area divided by gradient.

These are computed using the formulae in [Quinn et al. 1991](https://doi.org/10.1002/hyp.3360050106).

The upstream area is computed by routing down slope with the fraction of the
area being routed to the next downstream pixel being proportional to the
gradient times the contour length. 

The local value of the gradient is computed using the average of a
subset of between pixel gradients. For a normal 'hill slope' cell these are the
gradients to downslope pixels weighted by contour length. In the case of
pixels which contain river channels the average of the gradients from upslope
pixels weighted by contour length us used.

These properties are computed in an algorithm that passes over
the data once in descending height. It is called as follows

```{r, calc_atb}
ctch$compute_properties()
```

The plot of the topographic index shows a pattern of increasing values closer
to the river channels

```{r, plot_atb}
## plot of topographic index (log(a/tan b))
ctch$plot_layer('atb')
```

<!-- Although not used in ordering the HRUs `dynatop` also provides the ability to compute flow distances for the hill slope cells. -->
<!-- The calculation of three distances is supported -->

<!-- - *shortest flow length* - the shortest length based on the pixel flow paths -->
<!-- to a channel -->
<!-- - *Dominant flow length* - the distance to a channel moving in the dominant -->
<!-- (largest fraction) flow direction from any grid cell -->
<!-- - *Expected flow length* - the distance to the channel based on a weighted -->
<!-- average of the down-slope flow lengths. Weights are given by the fraction of -->
<!-- flow in each direction. -->

<!-- The computation, in this example for the shortest flow length, is initiated with -->
<!-- ```{r, flow_length} -->
<!-- ctch$compute_flow_lengths(flow_routing="shortest") -->
<!-- ``` -->

The additional layers can be examined as expected
```{r, flow_length_plot}
ctch$get_layer()
```


## Adding additional layer

Properties may come in additional GIS layers. To demonstrate the addition of an
additional layer we will extract the filled dem

```{r ,extract_filled}
tmp <- ctch$get_layer("filled_dem")
```

then separate it into a layers representing land above and below 500m.

```{r height layer}
## T
tmp <- terra::ifel(tmp<=500,0,1)
```

The resulting raster object can now be added to the project with

```{r, add_height_layer}
ctch$add_layer(tmp, "greater_500")
ctch$get_layer()
```

# Classifying into Hydrological Response Units

Methods are provided for the classification of the catchment areas of similar
 hydrological response. The classifications generated in this process are
 augmented with a further distance based separation when generating a
 `dynatop` model (see following section).

By definition each channel length is treated as belonging to a single class. 

To classify the hillslope two methods can be used.

The `classify` method of a `dynatopGIS` allows a landscape property to be
*cut* into classes. 

For example to cut the topographic index for Swindale into 21 classes:

```{r, atb_split}
ctch$add_layer( ctch$classify("atb_20","atb",cuts=20) )
ctch$plot_layer("atb_20")
```

Providing a single value to the cuts argument determines the number of classes.
<!-- The values used to cut the variable can be extracted from the meta data with -->

<!-- ```{r,atb_splt_get_class} -->
<!-- ctch$get_method("atb_20") -->
<!-- ``` -->

The `combine_classes` method of a `dynatopGIS` allows classes to be combined
in two ways, which are applied in the order shown

- *pairing* - where unique combinations of classes create one new class
- *burning* - where a single class is imposed upon an area

To demonstrate a pairing combination consider combining the atb classes
generated above with the classification provided by the distance band

```{r, atb_20_band}
ctch$add_layer( ctch$combine_classes("atb_20_band",c("atb_20","band")) )
ctch$plot_layer("atb_20_band")
```

Additionally the land greater then 500 in altitude can be burnt in with

```{r, atb_20_band_burn}
terra::plot( ctch$combine_classes("atb_20_band_500",pairs=c("atb_20","band"),burns="greater_500") )
```

<!-- The each class in the combined classification the values of the classes used -->
<!-- in the computations can be returned -->


<!-- ```{r see_class} -->
<!-- head( ctch$get_method("atb_20_band_500")$groups ) -->
<!-- ``` -->

** IS THIS CORRECT?? **
Note that by giving the area to be burnt in a negative value when it was
generated above we have ensured that the values do not clash with those generated by the cuts which
(except potentially when a cut is NA) which will always be positive.

# Generating a dynamic TOPMODEL

A Dynamic TOPMODEL suitable for use with the ```dynatop``` package can be
generated using the `create_model` method. This uses an existing
classification 
to generate the model. The required model structure is
given in the vignettes of `dynatop` package and is not described here in
details. 

Since `dynatop` simulations make use of ordered HRUs to work downslope, a
classification which used a distance layer (see earlier section) which represents the ordered downslope
sequencing of the pixels is recommended.

Even if a distance layer is
not used in the classification one must be given to the `create_model`
method, so the resulting HRUs can be ordered.  

**Currently only the 'band' distance metric as used below will produce valid model**.

For example, in the case of the division of Swindale by topographic index into
21 classes and the bands directly the resulting model can be generated by

```{r, model_atb_split}
ctch$create_model(file.path(demo_dir,"new_model"),"atb_20")
```

Looking at the files within the `demo_dir` folder

```{r, model files}
list.files(demo_dir,pattern="new_model*")
```

shows that an addition raster map of the HRUs has been created in
`new_model.tif` along with a file `new_model.rds` containing a model
suitable for `dynatop`. The values on the map correspond to the `ìd` of the HRUs in
the `dynatop` model.

# Running a model

The purpose of this vignette is to provide an outline of the steps needed to
perform a Dynamic TOPMODEL simulation and introduce the formats of the data
input and returned. 

The data used in this example comes from Swindale and is contained within the package and can be loaded with
```{r, setup}
#library(dynatop)
data("Swindale")
```
which returns a variable `Swindale` with the following components:
```{r data_loaded}
#names(Swindale)
```

For better comparison with a likely analysis we separate these into a model
and observed data variables

```{r sep}
swindale_model <- readRDS(file.path(demo_dir,"new_model.rds")) #Swindale$model
swindale_obs <- Swindale$obs
```

# The model structure
A dynamic TOPMODEL is described in a list object. The list has the following
elements
```{r model_parts}
names(swindale_model)
```
which are described in [associated
vignette](The_Model_Object.html). The
[dynatopGIS](https://waternumbers.github.io/dynatopGIS/) package
can be used for constructing models. 

# Setting map locations

While not required for simulations if the location of a file containing
the a raster map of the locations of the HRUs is provided the states can be visualised within
dynatop.

The locations of the files are set in the `map` element of the model. For this
example the maps are located within the `extdata` directory of the package and
can be set using commands

```{r set_map}
swindale_model$map <- file.path(demo_dir,"new_model.tif")
##system.file("extdata","mdl","Swindale.tif",package="dynatop",mustWork=TRUE)
```

# Preparing input data

The input to the model is expected to take the form of an ```xts``` object with
constant time step whose column names are found in the 'precip' and 'pet'
variables of the HRUs descriptions in the model. Helpful functions for creating and
manipulating ```xts``` objects can be found
[here](http://rstudio-pubs-static.s3.amazonaws.com/288218_117e183e74964557a5da4fc5902fc671.html),
see also the `resample_xts` function in this package.

The discharge, precipitation and potential evapotranspiration (PET) inputs for
Swindale are contained with `swindale_obs` on a 15 minute time step.
```{r, obs}
head(swindale_obs)
```
Note the discharge is in m$^{3}$/s while the precipitation and PET are in m
accumulated over the time step.

To use the data with the model we need to set the names of the time series
inputs within the model. In this case this is already done as can be seen by
inspecting the `precip` and `pet` values for the individual HRU's
```{r, set_obs_names}
head(swindale_model$hru[[1]]$precip)
head(swindale_model$hru[[1]]$pet)
```

# Altering parameters

The parameter values are stored within the HRU descriptions. 
Which parameters are present depends upon the options selected
for the transmissivity and channel solution. Details can be found in the
[HRU](HRU.html) Vignettes.

Altering parameter values requires changing their values in the HRU definitions. 
For this catchment all HRU have the same parameter values. For this simulation 
we change the parameter vectors to be more representative of the catchment
```{r, change_param}
hru <- swindale_model$hru
for(ii in 1:length(hru)){
    if(!("endNode" %in% names(hru[[ii]]$class))){
        ## then HRU is not a channel
        ## saturated zone parameters
        hru[[ii]]$sz$parameters["m"] <- 0.0063
        hru[[ii]]$sz$parameters["t_0"] <- exp(7.46)
        ## unsaturated zone parameters
        hru[[ii]]$uz$parameters["t_d"] <- 8*60*60
        ## root zone parameters
        hru[[ii]]$rz$parameters["s_rzmax"] <- 0.1
        ## surface parameters
        hru[[ii]]$sf$parameters["c_sf"] <- 0.4
    }else{
        ## then HRU is a channel - set so no subsurface response
        ## saturated zone parameters
        hru[[ii]]$sz$parameters["t_0"] <- 0
        ## root zone parameters
        hru[[ii]]$rz$parameters["s_rzmax"] <- 0.001
        ## surface parameters
        hru[[ii]]$sf$parameters["c_sf"] <- 0.8
    }
    ## initialisation parameters
    hru[[ii]]$initialisation["s_rz_0"] <- 0.98
    hru[[ii]]$initialisation["r_uz_sz_0"] <- 1.755582e-07 ## initial outflow divided by catchment area
}
```

# Creating the dynatop Object

Simulations are performed by embedding the model and the observed data into
a `dynatop` object. First the object is created using the model in list form

```{r create_object}
ctch_mdl <- dynatop$new(hru,map=swindale_model$map)
```

This step performs some basic checks on the model for consistency.
The data can then be added

```{r add_data}
ctch_mdl$add_data(swindale_obs)
```

# Running dynamic TOPMODEL

The model can be run with the `sim` method. The methods check that suitable
input data is available, but not how it was generated.

The initial states of the simulations can be specified in the model
object. If, as in the case of this example, the states are not specified then
any attempt to perform a simulation will fail.

```{r sim_fail, error=TRUE, purl=FALSE}
ctch_mdl$sim(swindale_model$output_flux)
```

Rather then editing each HRU to provide states they can be initialised using the `initialise` method which requires
an initial recharge rate to be set for each HRU. In the following we initialise the states and plot
the initial saturated zone storage deficit, using the chaining of commands.

```{r initialise}
ctch_mdl$initialise()$plot_state("s_sz")
```

The simulation can now be performed and the flow at the gauge extracted with

```{r sim1}
sim1 <- ctch_mdl$sim(swindale_model$output_flux)$get_output()
```

Note that the states of the system are now those at the end of the simulation
for example:

```{r new_states}
ctch_mdl$plot_state("s_sz")
```

Rerunning the simulation with the new initial states will of course produce
different results. Output for the above examples can be plotted against observed discharge for comparison as follows:

```{r sim2}
sim2 <- ctch_mdl$sim(swindale_model$output_flux)$get_output()
out <- merge( merge(swindale_obs,sim1),sim2)
names(out) <- c(names(swindale_obs),'sim_1','sim_2')
plot(out[,c('flow','sim_1','sim_2')], main="Discharge",ylab="m3/s",legend.loc="topright")
```

# Mass balance

It is possible to output the mass balance check for each time step of the
simulation using the `get_mass_errors` method. The returned `xts` object gives 
the volumes in the states at the start and end of the time step along with the
other fluxes as volumes. This can easily be used to plot the errors as shown below.

```{r mass_check}
mb <- ctch_mdl$get_mass_errors()
plot( mb[,6] , main="Mass Error", ylab="[m^3]")
```

