---
title: "The Model Object"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: false
    toc: true
    toc_depth: 2
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{The Model Object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The purpose of this vignette is to document the structure of a Dynamic
TOPMODEL model 'object'; that is the structure of the model that can be generated in
the [dynatopGIS](https://waternumbers.github.io/dynatopGIS/) package and is
used in calls to the ```dynatop``` function in the current package. Currently
the model is stored as an R variable of the list class.

This should be read alongside the ["Getting
Started"](dynatop.html) vignette which demonstrates the manipulation and use
of a model object.

The model structure will be documented with reference to the model in the test
catchment data supplied with the package. This can be loaded with

```{r, setup}
library(dynatop)
data("Swindale")
model <- Swindale$model
```

The top level of the model contains the following elements:

```{r model_parts}
names(model)
```

We consider each of these separately.

## map

The `map` element gives the path to the raster map of HRU locations which can
be used in the creation of a new ```dynatop``` object to enable plotting of
the state variables.

## output_flux

The `output_flux` element contains a data frame which defines the series to be output
during a simulation. This is passed to the `sim` method of the ```dynatop```
object. The columns of `output_flux` are given in Table
\@ref(tab:output-flux).

A single output value for each series is given for each data time step. If
multiple rows of the data frame contain the same name the values of the series are summed. If the simulation is run with substeps then the outputs are
averaged over the sub steps. Note that strictly only "q_sf", "q_sf_in", "q_sz"
and "q_sz_in" are fluxes, the remainder are volumes.

| Name    | Class   | Unit  | Description                                                                                                                              |
| ---     | ---     | ---   | -----                                                                                                                                    |
| name      | character | -   | name given to the output series |
| id    | integer | - | id of the HRU from which the output is to be taken |
| flux | character | -    | Flux to be recorded. One of "precip", "pet", "aet", "q_sf", "q_sf_in", "q_sz", "q_sz_in", "s_sf", "s_rz", "s_uz", "s_sz", "v_sf_rz", "v_rz_uz" or "v_uz_sz". |

Table: (\#tab:output-flux) Description of `output_flux` data frame columns

## hru

The `hru` contains a list, where each list element is the description of a
single HRU. Each HRU is described by a list containing the elements outlined
in Table \@ref(tab:model-table) and described in more detail in the sections
below. In many cases the naming of the variables within the vectors relates to
the variables names in the [HRU vignette](HRU.html) as given in
[implementation document](Notes_on_the_implementation_of_Dynamic_TOPMODEL.html).

| Name              | Class   | Description                                              |
|-------------------|---------|----------------------------------------------------------|
| id                | integer | The unique ID of the HRU                                 |
| states            | numeric | Named numeric vector of states                           |
| properties        | numeric | Named numeric vector of properties                       |
| sf                | list    | Description of the surface zone solution                 |
| rz                | list    | Description of the surface zone solution                 |
| uz                | list    | Description of the unsaturated zone solution             |
| sz                | list    | Description of the saturated solution                    |
| sf_flow_direction | list    | Description of the surface zone outflows to other HRUs   |
| sz_flow_direction | list    | Description of the saturated zone outflows to other HRUs |
| initialisation    | numeric | Named numeric vector of initialisation parameters        |
| precip            | list    | Description of the precipitation input to the HRU        |
| pet               | list    | Description of the PET input to the HRU                  |
| class             | list    | Classification information from dynatopGIS               |

Table: (\#tab:model-table) Description of top-level model components

### states

A numeric vector of states, see Table \@ref(tab:states), such as

```{r}
model$hru[[1]]$states
```

| Name | Class   | Unit | Description                                   |
|------|---------|------|-----------------------------------------------|
| s_sf | numeric | m    | Surface Zone storage volume per unit area     |
| s_rz | numeric | m    | Root Zone storage volume per unit area        |
| s_uz | numeric | m    | Unsaturated Zone storage volume per unit area |
| s_sz | numeric | m    | Saturated Zone storage deficit per unit area  |

Table: (\#tab:states) Description of numeric state vector

### properties

A numeric vector of properties, see Table \@ref(tab:prop) such as

```{r}
model$hru[[1]]$properties
```

| Name     | Class   | Unit  | Description                 |
|----------|---------|-------|-----------------------------|
| area     | numeric | m$^2$ | Surface area of the HRU     |
| width    | numeric | m     | Width of the HRU            |
| Dx       | numeric | m     | Effective length of the HRU |
| gradient | numeric | -     | Average gradient            |

Table: (\#tab:prop) Description of numeric state vector

### sf, rz, uz,& sz

The list variables describe the surface (sf), root zone (rz), unsaturated zone
(uz) ans saturated zone (sz) representations. Each list contains a character
vector `type` giving the type of representation and `parameters` a numeric
vector of names parameters. For example

```{r}
model$hru[[1]]$sz
```

The description of the types and parameters are given in the [HRU
vignette](HRU.html)

### sf_flow_direction & sz_flow_direction

Each of these lists contains two vectors as specified in Table
\@ref(tab:flow). In the special case that no outflow goes to another HRU, such
as at the outlet of the catchment, the description of the flow direction takes
the following form:

```{r}
model$hru[[1]]$sf_flow_direction
```
| Name | Class   | Unit | Description                                                                                    |
|------|---------|------|------------------------------------------------------------------------------------------------|
| id   | integer | -    | Vector of id values for the HRUs to which the flow is going. Must be lower then the current id |
| frc  | numeric | 0-1  | Fraction of the flow which goes from the corresponding id. Must sum to 1                       |

Table: (\#tab:flow) Description of flow direction list variables


### initialisation

The named numeric vector contains the values to use in the initialisation of
the HRU as given in Table \@ref(tab:init) and shown below.

```{r}
model$hru[[1]]$initialisation
```

| Name     | Class   | Unit | Description                                                                |
|----------|---------|------|----------------------------------------------------------------------------|
| s_rz0    | numeric | 0-1  | Initial root zone depth expressed as fraction of maximum depth (parameter) |
| r_uz_sz0 | numeric | m/s  | Initial recharge to the saturated zone per unit area (parameter)           |

Table: (\#tab:init) Description of initialisation values

### precip & pet

These list contain the description of the precipitation (precip) and PET
(pet) inputs used by the HRU. Each list takes the form of two vectors as given
in Table \@ref(tab:input) and shown below. From this the average input value
is calculated.

```{r}
model$hru[[1]]$precip
```

| Name | Class     | Unit | Description                                                                      |
|------|-----------|------|----------------------------------------------------------------------------------|
| name | character | -    | Character vector containing the names of the input series to use.                |
| frc  | numeric   | 0-1  | The fraction of the HRU area for which the input series is valid. Must sum to 1. |

Table: (\#tab:input) Description of precip and pet lists


### class

Class is a list variable. while not used in the simulation code it can be used
to store information that might be of use in, for example, determining which
parameter values to apply to the HRU (see the ["Getting Started"](dynatop.html) vignette for
an example).

Currently if the model is constructed using 
[dynatopGIS](https://waternumbers.github.io/dynatopGIS/) `class` is populated
based on the classification and channel data used in the model build process,
but the format of this is subject to change.

